# Issue 427: "`rescale()` has odd behavior at end points", or "`drawPalette()` should do the work of `rescale()` to return a vector of colors according to the input parameters"

When using `rescale()` to create a color vector, it returns floating point numbers, which are poor indices for assigning colors. For example, in the code below, which assigns 10 colors to a vector 100 points long, only the *highest* value will be assigned the extreme color (dark red), rather than according to a sensible choice of `breaks` given the range and the number of colors provided.

```{r}
library(oce)
set.seed(123)
x <- rnorm(100)
xscaled <- rescale(x, rlow=1, rhigh=10)
xscaled
col <- oceColorsJet(10)[xscaled]
```

This is mostly ok (or at least not visibly wrong) when there are a lot of colors (and data points), but for instances where one would desire a finite number of colors (say, to plot values by month), it makes it impossible to line up a palette (with breaks) produced using `drawPalette()` and the vector of colors produced using `rescale()`.

Another example: plot `x` from above with breaks `seq(-4, 4, 2)`, and make a palette that matches the colors assigned to the points:

```{r}
breaks <- seq(-4, 4, 2)
nbreaks <- length(breaks)
ncol <- nbreaks - 1  # need one less color than breaks
col <- oceColorsJet(ncol)[rescale(x, min(breaks), max(breaks), 1, ncol)]
par(mar=c(3, 3, 1, 1))
drawPalette(breaks, col=oceColorsJet, breaks=breaks)
plot(x, col=col, pch=19, ylim=range(breaks))
abline(h=breaks)
```

It can clearly be seen that the colors are being assigned to the points incorrectly.

The issue outlined above occurs because `rescale()` returns floating point numbers, which when used as indices for resorting the palette vector are converted to integers. Thus, as shown, only the highest value corresponds to the `rhigh` arg, and thus is the only point to receive the upper extreme of the color range.

## Possible functional solution with drawPalette()??

A possible solution could be to make `drawPalette()` return a vector of colors that could be used to plot points, based on the `zlim` arg which it is given (i.e. could be the vector `x`) and the breaks specified

```{r eval=FALSE}
col <- drawPalette(x, col=oceColorsJet, breaks=seq(-4, 4, 2))
plot(x, col=col)
```

For this example note that the vector `col` would be of the appropriate length according to the provided `x` and the colors assigned properly according to the specified `breaks`.

