# Introduction

I'm unable to predict the eclipse that's coming up on the 15th, nor can I get
the one from 1970 that was in Carly Simon's song, although the latter used to
work (in OAR).


# Methods/results

Since I don't have my textbooks with me, I'll try using an online calculator to
get sun and moon angles and check that way.  (Probably ok.)

Look for Halifax (44.8544N, 63.1992W).  At time of test (April), add 3 hours to
local time to get UTC.

## Sunrise/sunset nominally 2014 April 1 in Halifax NS

A
[calculator](http://www.timeanddate.com/worldclock/astronomy.html?n=286&month=4&year=2014&obj=sun&afl=-1&day=1)
yields sunrise/set at times tested with vertical lines in ``423A.R``.

## Sun test

```{r}
source("423A.R")
```

## Moon test

A [calculator](http://www.timeanddate.com/worldclock/astronomy.html?n=286&month=4&year=2014&obj=moon&afl=-1&day=1)
yields moonrise/set at times tested with vertical lines in ``423B.R``.

```{r}
source("423B.R")
```

# Observations

Clearly something is wrong with the moon calculations.

# Conclusions

None yet.  May need to look at the book from which I got the formulae, for some
test cases that will let me narrow down on the error.

