# Introduction

There was a bug in moonAngle, which is now fixed, as shown below (be sure to build this from Rmd)


# Methods/results

Since I don't have my textbooks with me, I'll try using an online calculator to
get sun and moon angles and check that way.  (Probably ok.)

Look for Halifax (44.8544N, 63.1992W).  At time of test (April), add 3 hours to
local time to get UTC.

## Sunrise/sunset nominally 2014 April 1 in Halifax NS

A
[calculator](http://www.timeanddate.com/worldclock/astronomy.html?n=286&month=4&year=2014&obj=sun&afl=-1&day=1)
yields sunrise/set at times tested with vertical lines in ``423A.R``.

## Sun test

```{r}
source("423A.R")
```

## Moon test

A [calculator](http://www.timeanddate.com/worldclock/astronomy.html?n=286&month=4&year=2014&obj=moon&afl=-1&day=1)
yields moonrise/set at times tested with vertical lines in ``423B.R``.

```{r}
source("423B.R")
```

# Observations

Looks fine, on this test to withing 1 degree on azimuth and 1 minute of time.


# Conclusions

There was a bug in switching lon for lat, which is now fixed in the ``develop``
branch, after commit f2f7f99c79a6286fd6bd0d638e991e2668b51c16.

